CS 51 Final Project Report

### 2013-05-06 DraftSpec_Annotated.tex
adding stuff to specs
fixed specs

\begin{document}
\title{CS 51 Final Project Functional Specification{\annot{:Matrix Module and Simplex Algorithm}}
---{\annot{Annotated!}}}
\author{
Luis Perez $|$ luisperez@college.harvard.edu \\ 
Andy Shi $|$ andyshi@college.harvard.edu \\ 
Zihao Wang $|$ zihaowang01@college.harvard.edu \\ 
Ding Zhou $|$ dzhou@college.harvard.edu
}
\date{May 5, 2013}
\maketitle

@@

The row reduction algorithm will certainly be one of the most fundamental
algorithms in our matrix library implementation. From our current understanding
of simplex, it seems like it will heavily rely on row reduction {\annot{The
simplex algorithm ended up relying on a different type of row reduction, so we
couldn't just plug in a matrix into our row reduction algorithm. We had to do a
lot of extra work to get simplex. Though, many of the helper functions for row
reduction ended up being extremely useful for Simplex, so this was non as pointless
as some might thing}}. The only algorithm we can come up with so
far is Gauss-Jordan, though we have found some algorithms which are more suited
for computation by computers with imprecise storage of floats {\annot{In fact,
we just had to be careful which element to choose as a pivot. Futhermore, we
ended up implementing arbitrary precision, so this became a non-issue}}.

>>>>

\begin{document}
\title{CS 51 Final Project Functional Specification{\annot{Matrix Module and Simplex Algorithm}}
---{\annot{Annotated!}}}
\author{
Luis Perez $|$ luisperez@college.harvard.edu \\ 
Andy Shi $|$ andyshi@college.harvard.edu \\ 
Zihao Wang $|$ zihaowang01@college.harvard.edu \\ 
Ding Zhou $|$ dzhou@college.harvard.edu
}
\date{May 5, 2013}
\maketitle

@@

The row reduction algorithm will certainly be one of the most fundamental
algorithms in our matrix library implementation. From our current understanding
of simplex, it seems like it will heavily rely on row reduction {\annot{The
simplex algorithm ended up relying on a different type of row reduction, so we
couldn't just plug in a matrix into our row reduction algorithm. We had to do a
lot of extra work to get simplex. Though, many of the helper functions for row
reduction ended up being extremely useful for Simplex, so this was not as
pointless as some might think}}. The only algorithm we can come up with so far
is Gauss-Jordan, though we have found some algorithms which are more suited for
computation by computers with imprecise storage of floats {\annot{In fact, we
just had to be careful which element to choose as a pivot. Futhermore, we ended
up implementing arbitrary precision, via Ocaml's num library,  so this became a
non-issue}}.

### 2013-05-06 report.tex
fixed specs

\documentclass[letterpaper,11pt]{article}
% Change the header if you're going to change this.

@@

@@

@@

In the real world, people love to make money. Even in virtual worlds (bitcoin?),
people love to make money. Basically, many problems in business and even
everyday life (but especially business), boil down to maximizing profit or
minimizing costs. Of course, when you speak of minimization or maximization
there are always constraints. Resources are finite and time is precious.
Therefore, in an abstract sense, the problem boils down to minimizing or
maximizing some quantity under certain constraints. For example, you might want
to maximize the number of textbooks x,y,z you buy under the constraints that you
can't buy more than five from one company, the total cost can't be more
than \$500, and the last two books you buy must cost more than \$75. This problem
is precisely the type of problems that our program can now solve. How did we do
this?

To start off, we created a framework. We implemented a matrix module with
standard matrix operations such as multiplication, addition, map (where scaling
follows) as well as many extra features such as row reduction, determinant,
trace, and many more. The matrix module basically allows a user to do anything
imaginable with a matrix, in addition to providing read and write functionality
for easy loading of data, and printing functionality for straightforward viewing
of the abstract matrix. However the intellectually challenging part of our
project is the implementation of the simplex algorithm. We needed to create a
matrix module simply because OCaml doesn't provide one. We use many of the
functions in the matrix module to create and manipulate simplex systems within
our simplex module. This allowed us to abstract away yet another part of our
project.

The simplex algorithm is a type of linear programming that allows one to
optimize a linear equation given linear constraints (like the example problem
at the beginning). In mathematical terms, the simplex program operates on
linear programs in standard form. That is, it seeks to minimize (or maximize)
c.x subject to $A.x = b$ (Wikipedia). Let's consider the inequalities: $30x +
15y + 18z \geq 500, x \geq 5,y \geq 5, z \geq 5$, and $15y+18z \geq 75$ and the objective
functions $x +y + z$ where $x, y, z \geq 0$. As you will notice, this is precisely
the system of equations necessary to solve our book example above. The
algorithm converts the inequalities into equalities by adding positive slack
variables. Then it represents this set of equations in the matrix form where
the coefficients of the terms we maximize are on the first row and each column
represents a variable. This provides affordable storage of most of the
information necessary to solve the linear program. Furthermore, we then define
basic and non-basic variables (kept track of in a list), which together with
the aforementioned matrix, create a simplex system.

The first function necessary for simplex is called \verb@simple_solve@. This
function, in abstract terms, finds the minimum value of a passed in system (as
described above). When we call the function \verb@simple_solve@, the program
finds the entering and leaving variables which are the variable that would
exchange position from basic and non-basic (slack to non-slack in the initial
case). Here, the algorithm is making the assumption that the system has a
feasible solution where all of our non-basic variables are zero, and therefore
exchanges the role of one of these variables with the corresponding basic
variable. In geometric terms, we are moving along one edge of the n-gon formed
by the constraints and attempting to minimize. This step is programmatically
accomplished by the helper function pivot. The function \verb@simple_solve@
continues to loop (moving along edges and minimizing) until the first
row---which represents the term we want to optimize consists of only nonpositive
values or there is no possible entering variable (ie, any change in the
variables will only increase our answer, not minimize it). In the first case,
algebraically, it means that the constraint can only increase if the variables
change. Therefore, we have found our minimum value which is the constant stored
at the end of the first row. This is the optimized value of the objective
function. In the second case, where no suitable entering variable can be found,
it means that the system can be decreased infinitely as pleased and therefore it
is Unbounded. In the case where a solution exists, one possible value for the
original variables can also be found by looking at their corresponding columns
and checking to see if they are basic. If there are an infinite number of these,
our algorithm returns only one.

What if we had the constraint $x + y + z \geq -10$. As stated above, our
\verb@simple_solve@ makes the assumption that one of the feasible solutions is
the zero solution, but if we initially set $x, y$, and $z$ to zero the
constraint does not hold. We therefore have a helper function called
\verb@initialize_simplex@ that converts everything into the ``simple'' form so
we can just recursively call \verb@simple_solve@. It does this by following a
very complex algorithm better detailed below. Our final  simplex algorithm now
accounts for all cases as long as the equations are linear. It also handles not
only less than or equal constraints, but also greater than or equal to and
equality constraints.  Finally, not only can it minimize, but it can also
maximize. You can follow the algorithm at each step on www.simplexme.com, but we
found that SimplexMe only works for the ``simple'' case. It has errors for large
matrices and the infeasible and unbounded cases.

\section{Planning}

@@

Our matrix module is fairly robust and it certainly implements everything that
is necessary to program the simplex algorithm (which was our main goal for this
module). Yet, it goes further. We managed to implement determinant, trace,
row-reduction,inverse, LU decomposition, and more (all extra functions that we
classified under our ``cool features'' category. There were only minor
points that we did not achieve, such as the eigen function.

As to the Elts module, we really went far and beyond. Because we coded
everything from an abstract point of view, once we had completed our project we
were able to go back and add a new Nums module which provided arbitrary
precision and size (ie, implementing arbitrary float arithmetic and bignum
arithmetic). Of course, to be honest, we really didn't actually code most of
this up. We used the built in Nums library. But as can be seen inside of
Elts.ml, our code was planned out so well that we can replaced the foundation of
mathematics (\verb@ORDERED_AND_OPERATIONAL@) with only a few changes in code.

The Simplex module also went further than expected (though it certainly was more
difficult  than expected too). Not only did we implement the ability to solve
simple linear programs, but we can now solve with certainty any linear program
(dual and primal), and provide the use with not only the optimal value, but also
one possible solution (which, in most cases is the only solution).

@@

Testing the system also proved to be a problem in its own right. Even after we
got all of the parsing functions down and were certain that the data was being
loaded correctly, we were never really able to find a large set of simplex
example problems. In the end, we resorted to each team member individually
finding a problem and solution (with Mathematica, of course, since SimplexMe
proved unreliable), and then typing out the data into a text file so we could
test it. Once we had all the data, we decided to program an interface that would
allow for expanding the testability of the code. More information on this can be
found the the \verb@README@.

As to the design and implementation of our actual code base, it definitely can
be improved. It is the best it can be given the fact that we hadn't already done
this same problem before, but as with anything, it can be better. For example,
our \verb@initialize_simplex@ function ended up being very complex and
complicated. We all understand the general idea of what it does, but not all of
us could implement it by ourselves from scratch. Basically,
\verb@initialize_simplex@ takes in a system and adds the slack variables to it.
During this step, if the system appears unfeasible, it adds an extra variable
and attempts to minimize that (ie, replaces the objective function). If this new
system has a minimum of zero, then the original system is feasible and all we
need to do is substitute the old objective back in appropriately. Otherwise, the
original system is unfeasible and the program returns None. In the first case,
an extra pivot operation might be necessary, but again, initialize simplex takes
care of this.  In the end, it calls \verb@simple_solve@ (the function we
originally wrote) on this now simple case and a solution is produced. However,
we did run our simplex algorithm against an equation with 100 variables and 100
constraints (\verb@test22.txt@), and it solved it in under 1 second. Even though
the worst case running time of simplex is not polynomial (it's exponential), on
most real world cases simplex runs very quickly, as evidenced by our test.

Each member of the team contributed to the project. The work was split between
the matrix and elts modules, parsing the data, and the simplex algorithm. Andy
and Luis implemented most of the matrix module and parsing function, along with
the general framework of the program that allowed us to efficiently run and test
our code. Luis, Jason, and Ding worked on implementing and debugging the simplex
algorithm. Ding focused on comprehending every step of the algorithm as
presented in the Algorithms book and then presenting that information in a more
comprehensive manner to the the group, while Jason focused on understanding the
process behind simplex and on coming up with example problems by hand. Jason and
Ding worked on testing simplex, and Luis and Andy provided much of the
functionality that allowed for easy testing.

@@

Start as early as possible and put some serious work in. Don't leave the video
until the last minute. This is a big project and showing it off is a very
rewarding part of it, especially when done well. Also try to understand the
general outline for every step before serious implementation, even if it means
delaying the implementation. Designing the framework is crucial and it is easier
to get it right the first time. Make sure to write test functions as you write
functions and to test each function right after you finish writing it. This way,
you will avoid having to hunt down a small bug in one obscure part of your code.
You'll know exactly when and where your code went wrong.

>>>>

\documentclass[letterpaper,12pt]{article}
% Change the header if you're going to change this.

@@

\newcommand{\vect}[1]{\mathbf{#1}}

@@

Link to demo video: \url{}

@@

In the real world, people love to make money. Even in virtual worlds (bitcoin?),
people love to make money. Basically, many problems in business and even
everyday life (but especially business), boil down to maximizing profit or
minimizing costs. Of course, when you speak of minimization or maximization
there are always constraints. Resources are finite and time is precious.
Therefore, in an abstract sense, the problem boils down to minimizing or
maximizing some quantity under certain constraints. For example, you might want
to maximize the number of textbooks $x,y,z$ you buy under the constraints that you
can't buy more than five from one company, the total cost can't be more
than \$500, and the last two books you buy must cost more than \$75. This problem
is precisely the type of problem that our program can now solve. How did we do
this?

To start off, we created a framework. We implemented a matrix module with
standard matrix operations such as multiplication, addition, map (where scaling
follows) as well as many extra features such as row reduction, determinant,
trace, and many more. The matrix module basically allows a user to do almost
anything imaginable with a matrix. In addition, we provided read and write
functionality for easy loading of data, and printing functionality for
straightforward viewing of the abstract matrix. However the intellectually
challenging part of our project is the implementation of the simplex algorithm.
We needed to create a matrix module simply because OCaml doesn't provide one
(also because we just wanted to implement a matrix module, of course). We use
many of the functions in the matrix module to create and manipulate simplex
systems within our simplex module. This allowed us to abstract away yet another
part of our project.

The simplex algorithm is a type of linear programming that allows one to
optimize a linear equation given linear constraints (like the example problem at
the beginning). In mathematical terms, the simplex program operates on linear
programs in standard form. That is, it seeks to minimize (or maximize)
$C\vect{x}$ subject to $A\vect{x} = \vect{b} $ (Wikipedia). $\vect{x}$ and
$\vect{b}$ are vectors. $x$ contains the variables in the equation we want to
optimize (the objective), and $b$ contains the constants in the constraints. $C$
and $A$ are matrices, and they represent the coefficients in the objective and
the constraints, respectively. Let's consider the inequalities: $30x + 15y + 18z
\geq 500, x \geq 5,y \geq 5, z \geq 5$, and $15y+18z \geq 75$ and the objective
functions $x +y + z$ where $x, y, z \geq 0$. As you will notice, this is
precisely the system of equations necessary to solve our book example above. The
algorithm converts the inequalities into equalities by adding positive slack
variables. Then it represents this set of equations in the matrix form where the
coefficients of the terms we maximize are on the first row and each column
represents a variable. This provides affordable storage of most of the
information necessary to solve the linear program.  Furthermore, we then define
basic and non-basic variables (kept track of in a list), which together with the
aforementioned matrix, create a simplex system.

The first function necessary for simplex is called \verb@simple_solve@. This
function, in abstract terms, finds the minimum value of a passed in system (as
described above). When we call the function \verb@simple_solve@, the program
finds the entering and leaving variables which are the variable that would
exchange position from basic and non-basic (slack to non-slack in the initial
case). Here, the algorithm is making the assumption that the system has a
feasible solution where all of our non-basic variables are zero, and therefore
exchanges the role of one of these variables with the corresponding basic
variable. In geometric terms, we are moving along one edge of the n-gon formed
by the constraints and attempting to minimize. This step is accomplished by the
helper function pivot. The function \verb@simple_solve@ continues to loop
(moving along edges and minimizing) until the first row---which represents the
term we want to optimize consists of only non-positive values or there is no
possible entering variable (ie, any change in the variables will only increase
our answer, not minimize it). In the first case, algebraically, it means that
the constraint can only increase if the variables change. Therefore, we have
found our minimum value which is the constant stored at the end of the first
row. This is the optimized value of the objective function. In the second case,
where no suitable entering variable can be found, it means that the system can
be decreased infinitely as pleased and therefore it is Unbounded. In the case
where a solution exists, one possible value for the original variables can also
be found by looking at their corresponding columns and checking to see if they
are basic. If there are an infinite number of these, our algorithm returns only
one.

What if we had the constraint $x + y + z \geq -10$? As stated above, our
\verb@simple_solve@ makes the assumption that one of the feasible solutions is
the zero solution, but if we initially set $x, y$, and $z$ to 0 (our algorithm
generates a ``basic solution'' by doing this step) the constraint does not hold.
We therefore have a helper function called \verb@initialize_simplex@ that
converts everything into the ``simple'' form so we can just recursively call
\verb@simple_solve@. It does this by following a very complex algorithm better
detailed below. Our final  simplex algorithm now accounts for all cases as long
as the equations are linear. It also handles not only less than or equal
constraints, but also greater than or equal to and equality constraints.
Finally, not only can it minimize, but it can also maximize. You can follow the
algorithm at each step on www.simplexme.com (we used this to debug parts of our
code), but we found that SimplexMe only works for the ``simple'' case. It has
errors for large matrices and the infeasible and unbounded cases.

\section{Planning}
Here is our \href{./DraftSpec_Annotated.pdf}{functional spec} and our
\href{./FinalSpec_Annotated.pdf}{technical spec} (the links open new PDFs).

@@

Our matrix module is fairly robust and it certainly implements everything that
is necessary to program the simplex algorithm (which was our main goal for this
module). Yet, it goes further. We managed to implement determinant, trace,
row-reduction,inverse, LU decomposition, and more (all extra functions that we
classified under our ``cool features'' category. There were only minor points
that we did not achieve, such as the eigen function, which finds eigenvectors
and eigenvalues.

As to the Elts module, we really went far and beyond. Because we coded
everything from an abstract point of view, once we had completed our project we
were able to go back and add a new Nums module which provided arbitrary
precision and size (ie, implementing arbitrary float arithmetic and bignum
arithmetic). Of course, to be honest, we really didn't actually code most of
this up. We used the built in Nums library. But as can be seen inside of
Elts.ml, our code was planned out so well that we can replace the foundation of
the mathematics in our program (\verb@ORDERED_AND_OPERATIONAL@) with only a few
changes in code.

The Simplex module also went further than expected (though it certainly was more
difficult to implement than we expected). Not only did we implement the ability
to solve simple linear programs, but we can now solve with certainty any linear
program (dual and primal---dual requires \verb@initialize_simplex@ while primal
is the ``simple'' case), and provide the user with not only the optimal value,
but also one possible solution (which, in most cases is the only solution).

@@

Testing the system also proved to be a problem in its own right. Even after we
got all of the parsing functions down and were certain that the data was being
loaded correctly, we were never really able to find a large set of simplex
example problems. In the end, we resorted to each team member individually
finding a problem and solution (with Mathematica, of course, since SimplexMe
proved unreliable), and then typing out the data into a text file so we could
test it. Once we had all the data, we decided to program an interface that would
allow for expanding the testability of the code. More information on this can be
found in the \verb@README@.

To use the simplex algorithm, we first parse a standardized plain text file
containing the parameters of the objective and the constraints. The first line
of this file must either be \verb@max@ or \verb@min@, signalling a maximization
or minimization step. The next line denotes the objective function. The
coefficients of each of the variables in the objective are separated by columns
(decimals must be represented by fractions because we are using Ocaml's num
library). The last element in this is the ``constant'' (5 if the objective is
$2x + 3y + 5$). The next line must read \verb@subject to@ to help us with
parsing. The next lines are the constraints, where the coefficient of each
variable that appears in the objective must appear (in the same order as in the
objective), separated by commas. We also allow the user to enter if the
constraint is a $\geq, \leq$, or $=$ condition. Again, the last element is the
constant (5 in the case of $x + y \leq 5$). It is assumed that all the variables
in the objective function are $\geq 0$.

Our program will parse this text file and generate a matrix representing the
objective and the constraints. Additionally, our program adds slack variables
(for example, (we change the inequalities in the constraints to equations---$3x
+ 2y \leq 5$ becomes $3x + 2y + s = 5$, where $s \geq 0$). This makes the
problem easier to solve).

The simplex algorithm basically takes the objective function and attempts to
find a ``basic solution'' which will satisfy all the constraints. Even if this
solution is trivial (like setting all the variables in the objective to 0), we
take this solution and keep optimizing it by replacing a variable not in the
objective function with a variable that is. We check the objective to make sure
none of the coefficients are negative; if they are, we are finished and we have
the optimum value in the top right corner of the matrix (because we can set all
the variables in the objective to 0 and not violate the constraints). Our
``simple'' case is the case where we want to minimize the objective and all the
constraints are $\leq$ relations. However, we can change any problem to this
form. Maximization can be achieved by minimizing the negative of the objective.
Similarly, a $\geq$ constraint can be changed to a $\leq$ constraint by
multiplying each side of the inequality by $-1$. Finally, an $=$ constraint can
be changed into a $\geq$ and a $\leq$.

As to the design and implementation of our actual code base, it definitely can
be improved. It is the best it can be given the fact that we hadn't already done
this same problem before, but as with anything, it can be better. For example,
our \verb@initialize_simplex@ function ended up being very complex and
complicated. We all understand the general idea of what it does, but not all of
us could implement it by ourselves from scratch. Basically,
\verb@initialize_simplex@ takes in a system and adds the slack variables to it.  
During this step, if the algorithm determines a system (the objective
function and the constraints) to be unfeasible, it adds an extra variable and
attempts to minimize that (ie, replaces the objective function). If this new
system has a minimum of 0, then the original system is feasible and all we need
to do is substitute the old objective back in appropriately. Otherwise, the
original system is unfeasible and the program returns \verb@None@. In the first
case, an extra pivot operation might be necessary, but again,
\verb@initialize_simplex@ takes care of this. In the end, it calls
\verb@simple_solve@ (the function we originally wrote) on this now simple case
and a solution is produced. However, we did run our simplex algorithm against an
equation with 100 variables and 100 constraints (\verb@test22.txt@), and it
solved it in under 1 second. Even though the worst case running time of simplex
is not polynomial (it's exponential), on most real world cases simplex runs very
quickly, as evidenced by our test.

Each member of the team contributed to the project. The work was split between
the matrix and elts modules, parsing the data, and the simplex algorithm. Andy
and Luis implemented most of the matrix module and parsing function, along with
the general framework of the program that allowed us to efficiently run and test
our code. Luis, Jason, and Ding worked on implementing and debugging the simplex
algorithm. Ding focused on comprehending every step of the algorithm as
presented in the Algorithms book and then presenting that information in a more
comprehensive manner to the group, while Jason focused on understanding the
process behind simplex and on coming up with example problems by hand. Jason and
Ding worked on testing simplex, and Luis and Andy provided much of the
functionality that allowed for easy testing.

@@

Start as early as possible and put some serious work in. Don't leave the video
until the last minute. This is a big project and showing it off is a very
rewarding part of it, especially when done well. Also try to understand the
general outline for every step before serious implementation, even if it means
delaying the implementation. Designing the framework is crucial and it makes the
whole process easier to get right the first time. Make sure to write test
functions as you write functions and to test each function right after you
finish writing it. This way, you will avoid having to hunt down a small bug in
one obscure part of your code.  You'll know exactly when and where your code
went wrong.
